---
title: "Manuscript 3: A tensor-based approach for exploring the effects of county interconnectedness on place-based health"
format: 
  pdf:
    message: false
    warning: false
    echo: false
    error: false
---

## Introduction

Migration plays a crucial role in shaping the demographic, social, and health landscapes of the United States. County-to-county migration, in particular, not only reflects the interconnectedness of regions but actively influences it, impacting local economies, social structures, and health outcomes in significant ways. Despite growing interest in understanding the health effects of migration, current analytical approaches struggle to capture the full complexity of migration flows and their multifaceted relationships with public health. Existing models often examine single or limited dimensions, missing the nuanced, multidimensional patterns essential for a comprehensive understanding of migration's health impacts. This paper proposes tensor factorization as a novel method for assessing the association between regional interconnectedness—driven by migration flows—and county-level health outcomes, specifically premature age-adjusted mortality rates.

County-level migration represents an important level of analysis due to its central role in influencing local and regional dynamics. Patterns of migration at this level affect and reflect health outcomes through various pathways, including economic conditions, population demographics, and access to social services. Premature age-adjusted mortality rates, the focal health measure in this analysis, provide a valuable proxy for the overall health and longevity of communities, helping to illuminate how migration may contribute to or mitigate health risks. However, capturing these interactions is challenging. Migration is inherently dynamic and occurs across complex, interwoven pathways that traditional methods struggle to quantify accurately. Commonly used approaches like Principal Component Analysis (PCA), regression models, and spatioeconomic models are limited in that they can only address a small number of dimensions simultaneously. PCA is primarily useful for data reduction but generally captures only bivariate relationships, while regression models are effective for assessing specific factors, such as the association between migration and health outcomes, but often fall short in capturing the complexity of migration flows and their interactions over time and space. Although spatioeconomic models incorporate geographic dimensions, they also face challenges in addressing the multifaceted nature of migration flows. Quadratic Assignment Procedure (QAP), another approach commonly applied in network analyses, can test the significance of relationships between networks by accounting for dependencies in network data [@liu2019]; however, QAP still lacks the multidimensional capacity necessary to explore migration's complex, interwoven patterns with health outcomes.

In response to these limitations, this study explores tensor factorization as a multidimensional approach that captures relationships between migration flows and health outcomes across multiple dimensions—origin, destination, and time—simultaneously. Tensor factorization, unlike traditional methods, is adept at separating data into interpretable patterns that reveal underlying connections among diverse variables. One way to conceptualize tensor factorization is by comparing it to baking a cake with different layers. Imagine that a dataset represents all the possible ingredients for creating a frosted cake, where one dimension holds the ingredients for the cake itself, another holds the ingredients for the frosting, and a third dimension holds different cake pan sizes. Tensor decomposition allows us to break this complex data into meaningful combinations: a classic chocolate cake, a simple vanilla cupcake, or even a chaotic blend of raw dough and frosting. This metaphor demonstrates how tensor factorization can uncover hidden patterns across dimensions—patterns that would remain obscured in a traditional analysis.

Tensor factorization has been applied successfully across numerous fields, demonstrating its versatility in uncovering hidden relationships. For example, it powers recommendation systems, such as Netflix’s algorithm, by simultaneously analyzing user preferences, item attributes, and ratings to uncover multidimensional patterns [@hidasi2012]. In image processing, tensor decomposition is employed to identify shared features across images, improving recognition and categorization algorithms. Public health researchers have also adopted tensor factorization to isolate seasonal trends in infectious diseases [@korevaar2020] and assess mental health impacts during the COVID-19 pandemic [@zhu2023]. Across these domains, tensor factorization has proven uniquely effective at illuminating complex relationships within multidimensional data—relationships that are often missed by other methods.

Researchers have also used tensor factorization to study human mobility patterns, further illustrating its relevance to migration and health research. For instance, one study employed tensor decomposition with geolocated Twitter data to analyze migration flows between European countries, identifying distinct migration types, such as student migration, long-term migration, and tourism [@nguyen2017]. Another study used GPS taxi data in Beijing to uncover both temporal movement patterns and "spatial communities," which researchers then linked to neighborhood stability [@wang2020]. In Singapore, tensor factorization was applied to public transportation data, enabling researchers to classify passenger types and journey patterns [@sun2016]. Moreover, a recent study in Liège, Belgium, used tensor decomposition to examine commuting patterns with mobile phone data, validating its potential for understanding population mobility [@kolda2008]. Similarly, researchers found that tensor factorization outperformed other methods in analyzing population mobility in urban centers using cell phone data [@balasubramaniam2020]. In these and similar studies, tensor factorization has consistently expanded upon conventional approaches in capturing the intricate dynamics of human mobility within and between regions.

Migration is a powerful influence on regional health, with the potential to either alleviate or exacerbate disparities depending on contextual factors. Regions experiencing an influx of younger, healthier individuals often see improvements in workforce participation and health outcomes, as these migrants contribute to economic stability and bring fresh skills and perspectives. Migration also enables individuals to leave areas with limited resources, potentially improving health outcomes in their new locations. However, migration can also deepen health disparities. Counties with high out-migration rates may lose younger, healthier populations, leading to an aging and more vulnerable remaining population that struggles with worsened health outcomes and strained local resources. Destination counties, particularly those unprepared for rapid population growth, may experience overcrowded health facilities and an increased demand for social services, which can stretch resources and exacerbate existing disparities.

In this paper, we aim to explore the utility of tensor factorization in uncovering complex and previously unexamined relationships between migration patterns and public health. By examining migration flows across origin, destination, and time dimensions, tensor factorization can offer a more comprehensive view of migration as a system that influences county-level health holistically, rather than isolating single effects. This approach provides an innovative, multidimensional perspective that can shed new light on the ways migration interacts with health across interconnected regions, offering valuable insights for further study.

## Methods

### Tensor Decomposition and Evaluation

#### Building the Tensor

We started by building a three-dimensional tensor from IRS county-to-county migration data, where the dimensions correspond to the **origin county**, **destination county**, and the **time period**. We use data representing 3109 origin and destination counties for years 2011 through 2019, with the IRS migration data capturing the number of individuals (based on exemptions claimed on tax returns) moving from each origin county to each destination county over a 12-month period (July 1 of the year prior to June 30 of the given year) [@soitax]. We replaced all "missing" migrations (ie no migration between a given county and destination) with zero such that our tensor contained all possible county-to-county migration patterns within the contiguous United States. Additionally, within-county migration (where origin and destination are the same) was set to zero to ensure that only inter-county migration was captured in the tensor. To normalize the data in the tensor, we applied a transformation using $log(1 + n)$, where $n$ represents the number of migrating individuals. This normalization approach preserves zero values in the data and helps mitigate the influence of large migration counts, which is consistent with the approach used in other tensor decompositions of migration data [@gong2024]. Therefore, each entry in the tensor, $X(o_i,d_j,t_k)$, represents the $log(1+n)$ transformed number of migrating individuals from origin $i$ to destination $j$ in time period $k$.

#### Tensor Decomposition

Since our tensor consists only of zero and positive values, we employed **Non-Negative Tensor Factorization (NTF)** from the `nnTensor` package in R [@nnTensor] which ensures that all components remain non-negative for interpretability. Additionally, the NTF function utilizes the Canonical Polyadic (CP) factorization method, which is ideal for creating interpretable results from sparse data because it produces a sum of independent components, each aligned directly with a mode, allowing for straightforward interpretation of patterns along each dimension [@haliassos2022].

Mathematically, the decomposition of the tensor using NTF is represented as: $$𝑋≈ \sum_{r=1}^R{o_{ir}•d_{jr}•t_{kr}}$$

where $R$ is the number of components (rank) for the decomposition and $o_{ir}$, $d_{jr}$, and $t_{kr}$ represent the factor vectors corresponding to each dimension: origin $i$, destination $j$, and year $k$ of the tensor. Each $rth$ component corresponds to a different "migration system," and the loadings (e.g., $o_{ir}$) represent the contribution of a county or time period to that component. A high loading on a particular variable signifies its importance in the migration system. Therefore, it follows that if $o_{ir}=0$ we know that origin $i$ is not in the $rth$ migration system. On the other hand, if $o_{ir}$ is large, we can infer that the $rth$ migration system has lots of migration activity from origin 𝑖. This is a direct extension of the decomposition of a three-way tensor described in Almquist et al, 2021 [@almquist2021].

To evaluate the quality of each decomposition, we prioritized minimizing the reconstruction error, measured by the Frobenius norm of the residual tensor—the difference between the original tensor and its reconstructed form. Lower Frobenius norm values indicate better approximations of the original tensor. Since tensor decomposition is a non-convex optimization problem, it can lead to multiple local minima rather than a single global minimum. This non-convexity means that the optimization landscape contains multiple peaks and valleys, making it challenging to find the best overall solution [@adali2022]. For each number of components $r$, we tested specific iteration counts (100, 200, and 500) across multiple random seeds, as varying seeds help avoid local minima and reveal the most stable solutions.

To assess the reproducibility of the decompositions, we calculated the Factor Match Score (FMS), which quantifies the similarity between the factor matrices derived from split-half sampling. The original tensor was randomly divided into two halves, ensuring no overlap between the subsets. Each half was subjected to non-negative tensor factorization (NTF) using the same rank, initialization method, and number of iterations. The FMS quantifies the alignment of the factor matrices $o_{ir}, d_{jr}, t_{kr}$ across the two decompositions.

For each component $r$, the cosine similarity was computed for the corresponding columns of the factor matrices from the two splits. Cosine similarity is defined as:

$$
 \text{cosine similarity}(\mathbf{x}, \mathbf{y}) = \frac{\mathbf{x} \cdot \mathbf{y}}{\|\mathbf{x}\| \|\mathbf{y}\|}
$$

where $x$ and $y$ are the respective columns of the factor matrices. The $FMS$ for each component was calculated as the product of the cosine similarities of the $r$-th columns of the three factor matrices:

$$
\text{FMS}_r = \text{sim}_o \cdot \text{sim}_d \cdot \text{sim}_t
$$

The overall $FMS$ was obtained by averaging the component-wise $FMS$ values across all components:

$$
 \text{FMS} = \frac{1}{R} \sum_{r=1}^{R} \text{FMS}_r
$$

where $R$ is the number of components. The $FMS$ ranges from 0 to 1, with higher values indicating greater reproducibility of the factorization across splits. This metric was used to evaluate the stability and reliability of the tensor decomposition.

Before generating the elbow plot, we filtered the decompositions by selecting only those with the maximum FMS score for each unique number of components, ensuring that only the most reproducible solutions were considered. We then plotted the Frobenius norm against the number of components to examine the relationship between model complexity and reconstruction accuracy, aiming to identify an "elbow" where additional components yield diminishing error reduction.

After reviewing the elbow plot to select an optimal number of components and iterations that effectively captured the migration system while balancing error minimization and simplicity, we generated maps of the tensor’s origin and destination dimensions for each component, along with a plot for the temporal dimension. To address skewness in the raw loadings, we ranked each origin, destination, and time period’s relative contribution from 1 to the number of unique loadings per dimension, assigning equal ranks to ties. We then used these maps to identify which decomposition best represented internal US migration patterns, comparing maps of decompositions with various numbers of components for interpretability, using Almquist's finding that a decomposition with six components is optimal as a guide.

### Spatial Modeling and Connection to County-Level Health

To determine whether county-interconnectedness (as measured by the ranked loadings of our tensor decomposition) is related to county-level health, we created a spatial panel model with random effects for spatial and temporal autocorrelation.

**Predictor Variables:** $o_{ir}$ $d_{jr}$, $t_{kr}$\
We included the loadings for origin, destination, and time dimensions as predictor variables, capturing the influence of county-interconnectedness (as represented by the tensor components) on health outcomes. Each $o_{ir}$ represents the loading for a unique origin county $i$ of a decomposition of rank $r$; $d_{jr}$ represents the loading for a unique destination county $j$ of a decomposition with rank $r$, and $t_{kr}$ represents the loading for a unique year $k$ of a decomposition with rank $r$.

**Outcome Variable:** $y_{it}$\
We use county-level premature age-adjusted mortality rates, available from CDC WONDER, as a proxy for overall county-level wellbeing. We use premature age-adjusted mortality rates for the same counties and the same time periods as represented in our migration tensor.

**Modifier:**\
We included urbanicity as a categorical term to test for the significance of urbanicity beyond the predictor loadings. This term enabled us to examine how the relationship between migration and mortality may vary by urbanicity. Our urbanicity term is based on the six urbanicity categories available from National Center for Health Statistics and includes the categories Large Central Metro, Large Fringe Metro, Medium Metro, Small Metro, Micropolitan, and Noncore [@dd2012].

We fit the model using the `spml` function from the `splm` R package [@splm-3], ensuring robustness against spatial dependencies in the data. To capture spatial relationships, we generated a spatial weights matrix using Queen's contiguity, which defines neighbors as counties sharing either a border or a corner. Our model also incorporated random effects to address both spatial and temporal autocorrelation. Our model is as follows:

$$ y_{it} = \beta_0 + \sum_{r=1}^{R} \beta_{o_r} o_{ir} + \sum_{r=1}^{R} \beta_{d_r} d_{jr} + \sum_{r=1}^{R} \beta_{t_r} t_{kr} + \sum_{n=2}^{6} \gamma_n \text{Urbanicity}_n + \mu_{it} $$

where:

-   $R$ is the total number of components, selected based on the elbow plot

-   $\beta_{o_r}$, $\beta_{d_r}$, and $\beta_{t_r}$ are the coefficients for each component $r$ of the origin, destination, and time loadings, respectively.

-   $\gamma_n$ are the coefficients for urbanicity categories, with the most urban category (large central metro) as the reference <!--# not sure if i should expand here or not .... \gamma_{\text{LFM}} \text{Large Fringe Metro}_i + \gamma_{\text{MM}} \text{Medium Metro}_i + \gamma_{\text{SM}} \text{Small Metro}_i + \gamma_{\text{MIC}} \text{Micropolitan}_i + \gamma_{\text{R}} \text{Rural}_i + u_{it}  -->

-   $\mu_{it}$ represents the spatial error term accounting for spatial and temporal autocorrelation in the data

## Results

### Decomposition Selection

Using the elbow plot below, we see that, as expected, the Frobenius norm of the residual generally decreased as the number of components increased, reflecting improved model fit with more complex models. However, we observed no clear relationship between the number of iterations and error, indicating that increasing iterations did not consistently improve the quality of the decomposition. This plot includes all iterations for each decomposition, with each dot representing a unique run with a different random seed, highlighting the variability introduced by initialization. Bolded points on the plot represent the decompositions that achieved the maximum (best) FMS score and the minimum error for each unique number of components, ensuring that the most stable and optimal solutions were emphasized. We observe some leveling-off of error (i.e., an elbow point) at around 3 components; however, we selected a decomposition with 6 components, as it had the lowest error and the highest FMS score. By systematically evaluating the decomposition across a range of complexities and visualizing the results, we aimed to confidently select the most interpretable decomposition.

```{r load the decomp and make the elbow plot}

load("data_processed_aim3/bestrun.RData") #this is called bestrun 

# for interpreting the "bestrun" output 

library(ggplot2)
library(dplyr)

# Filter the dataset to keep only rows where fms_score = 1 (highest reproducibility)
filtered_data <- bestrun %>%
  group_by(rank) %>%
  filter(fms_score == max(fms_score)) %>% 
  filter(error == min(error)) %>% 
  ungroup()  # Ungroup to avoid issues with further operations


# Create a scatter plot of error vs iterations, colored by rank and faceted by seed
ggplot(bestrun, aes(x = rank, y = error, color = as.factor(iterations))) +
  geom_point(size = 3) +
 # geom_line(aes(group = rank), linetype = "dashed") +
 # facet_wrap(~seed, scales = "free") +
  geom_point(data = filtered_data, aes(x = rank, y = error), 
             size = 5, shape = 21, fill = "black", color = "red") + # Add bolded points
  labs(
    title = "Selection of Best Rank, Iteration, and Seed",
    x = "Number of Components",
    y = "Reconstruction Error",
    color = "Iterations"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    legend.position = "bottom"
  )

# now pull the decomp with ncomp = 6 with fms=score = 1 
bestseed = bestrun %>% filter(rank == 6) %>% filter(fms_score == max(fms_score)) %>% select(seed)
bestiter = bestrun %>% filter(rank == 6) %>% filter(fms_score == max(fms_score)) %>% select(iterations)
bestpath = paste0("data_processed_aim3/ntf_model_rank6_seed", bestseed, "_iter", bestiter, ".RData")
load(bestpath)
#this is now called ntf_result

Orig6 <- ntf_result$A[[1]]  # Factor matrix for origin
Dest6 <- ntf_result$A[[2]]  # Factor matrix for destination
Time6 <- ntf_result$A[[3]]  # Factor matrix for year
```

### Interpretability of Migration Systems

After using the elbow plot to select a suitable number of components, we assessed the interpretability of the resulting decomposition. Unfortunately, we found that the decomposition was not interpretable due to several issues. The distribution of loadings was heavily skewed, with most counties having near-zero loadings and only a few counties dominating each component. This lack of meaningful variation made it difficult to identify distinct migration systems or draw meaningful conclusions about a county's role as an origin or destination within these systems. Additionally, the combinations of counties grouped together within components did not align with known migration patterns or plausible demographic or geographic relationships, further limiting interpretability. Density plots of the loadings, shown below, highlight this lack of distribution, with the majority of loadings clustered near zero. Maps showing the ranked loadings from the 6-component decomposition are provided in the appendix, but we note that these maps also reflect the limited interpretability due to the skewed and sparse distribution of loadings.

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Convert Orig6, Dest6, and Time6 matrices to data frames with a long format
Orig6_long <- Orig6 %>%
  as.data.frame() %>%
  mutate(Component = paste0("Component ", 1:6)) %>%
  pivot_longer(cols = -Component, names_to = "County", values_to = "Loading") %>%
  mutate(Dimension = "Origin")

Dest6_long <- Dest6 %>%
  as.data.frame() %>%
  mutate(Component = paste0("Component ", 1:6)) %>%
  pivot_longer(cols = -Component, names_to = "County", values_to = "Loading") %>%
  mutate(Dimension = "Destination")

Time6_long <- Time6 %>%
  as.data.frame() %>%
  mutate(Component = paste0("Component ", 1:6)) %>%
  pivot_longer(cols = -Component, names_to = "County", values_to = "Loading") %>%
  mutate(Dimension = "Time")

# Combine all dimensions into one data frame
loadings_long <- bind_rows(Orig6_long, Dest6_long, Time6_long)

# Generate density plots
ggplot(loadings_long, aes(x = Loading, color = Dimension, fill = Dimension)) +
  geom_density(alpha = 0.3, adjust = 1) +  # Transparent overlays for fill
  geom_density(size = 1, aes(linetype = Dimension)) +  # Add line types for clarity
  facet_wrap(~Component, scales = "free", ncol = 2) +
  labs(
    title = "Density Plots of Loadings Across Components and Dimensions",
    x = "Loading Value",
    y = "Density",
    color = "Dimension",
    fill = "Dimension",
    linetype = "Dimension"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold")
  )



```


### Relationship to County-level Health

Because we were unable to establish an interpretable tensor decomposition, we were unable to reliably model the relationship between a county's position within a migration system and county-level premature mortality rates. Without clearly defined components, it was not possible to assign counties to to migration systems, and subsequently link these migration systems to variations in premature mortality rates. This limitation highlights the challenges of applying tensor decomposition to datasets with many nodes and sparse connections, where the complexity and variability of migration flows may obscure patterns necessary for meaningful health analyses. An example model is provided in the appendix to illustrate what might be possible if a more interpretable decomposition were achieved.

<!--# f statistic? to test which components are significant ; likelihood ratio test - model with all predictors , model with some predictors - change in BIC if each predictor is dropped ; get relative importance of predictors to model -->

## Discussion

Through the exploratory use of tensor decomposition, we have attempted to identify and interpret migration systems within the context of internal migration in the US. Our analysis suggested six potential migration systems; however, they were not directly interpretable. While we explored possible associations between migration components and premature mortality rates, the results were inconclusive.  Despite this, we hope that the idea of examining a county's dual role as both an origin and a destination within migration systems - and its potential impact on health outcomes - will prove useful in future research. Understanding these roles could provide valuable insights into how migration-related factors contribute to health disparities and help inform targeted public health interventions. Future studies may benefit from alternative methods or enhanced preprocessing strategies to reduce sparsity and improve the interpretability of tensor components, thereby enabling a more robust exploration of the relationship between migration systems and health outcomes.

Taken together, these results suggest that the relationship between migration and health is multifaceted, with different migration patterns exerting varying effects on county-level mortality. The findings underscore the importance of considering both the origin and destination of migration flows, as well as the temporal dimension, when examining the public health implications of migration. This study underscores the exploratory nature of using tensor decomposition to examine the relationship between migration patterns and health outcomes. While it provides a novel framework for identifying and characterizing migration systems, further research is needed to establish clear and actionable connections between these systems and county-level health metrics.

An important challenge we encountered in this analysis lies in the instability and inconsistency of tensor decomposition methods. Tensor decomposition, while powerful for capturing multidimensional migration patterns, can often yield non-reproducible results due to its sensitivity to initial parameter settings, the choice of algorithm, and the nature of the data. In this study, we experimented with various initialization strategies to achieve a decomposition that best represented migration dynamics; however, even small variations in the initialization often produced substantially different decompositions. To address this, we used the Factor Match Score (FMS) based on cosine similarity to evaluate the stability of decomposition across split-half sampling. By quantifying the alignment of factor matrices between independent decompositions, we attempted to identify runs that were more consistent and reproducible. Despite this approach, achieving complete stability remained a challenge, highlighting the inherent variability in tensor decomposition methods.

This instability makes it difficult to interpret and replicate findings reliably across different studies or even within the same dataset. One key factor contributing to this inconsistency is the non-convex nature of the optimization involved in tensor decomposition [@adali2022]. Since multiple local minima exist, different runs can converge on distinct solutions that each capture different patterns in the data, even if the underlying data remains the same. This issue is particularly pronounced when using methods like Canonical Polyadic (CP) decomposition, where the model can fit the data well (i.e., minimizing error) but produce multiple different results with the same error that yield varying component interpretations across decompositions [@hale2021]. As a result, we observe contradictory migration patterns across different runs, which poses significant challenges in verifying the robustness of these findings.

Another layer of complexity arises from the limited interpretability of certain tensor components, which can shift depending on the data pre-processing steps and tensor rank selection. While methods such as non-negative tensor factorization (NTF) help improve interpretability by applying constraints to ensure positive values that align with real-world quantities, component selection remains subjective, relying heavily on domain expertise and anticipated outcomes. This subjectivity reduces reproducibility and weakens the potential of tensor decomposition as a consistent analytical tool in migration and health research. These challenges with instability underscore the need for future research to establish standardized protocols for validating tensor decomposition results [@adali2022]. Methods such as ensemble decomposition or consensus clustering may provide an avenue for reconciling diverse outputs generated by tensor decomposition to identify stable patterns.

The use of IRS migration data, which only includes individuals from household that file taxes, introduces a potential bias, as it excludes non-filers, such as low-income populations, undocumented migrants, and others who may not be captured by tax records. This could lead to an under-representation of certain demographic groups in the migration analysis, potentially skewing the results. That said, some researchers suggest that IRS data is actually more inclusive than ACS survey data since tax filings tend to have less non-response bias than survey data [@foster]. Additionally, if we observe a relationship between migration and health using the IRS data, this relationship is likely even stronger in reality as the excluded populations may experience even greater health disparities linked to migration. <!--# cite? -->

The IRS data itself has been subject to criticism for inconsistencies and volatility, particularly after 2011 when the IRS took over data preparation from the US Census Bureau. As DeWaard et al has noted, there are systemic issues with the IRS data that can lead to fluctuations in migration estimates, which are not fully explained by changes in IRS data collection methods [@dewaard2022]. This raises concerns about the reliability of the migration patterns observed in the post-2011 IRS data, which could affect the robustness of our findings.

Tensor decomposition has previously been applied to IRS migration data to analyze migration flows between U.S. metropolitan areas [@almquist2021]. This analysis involved a much less sparse tensor compared to our tensor, which includes all possible county-to-county migration flows for 3109 counties in the contiguous United States. With fewer nodes (metropolitan areas) and more migration events between them, tensor decomposition was better suited to the structure of metropolitan migration data. In contrast, the high sparsity of our tensor—arising from the large number of nodes and relatively rare migration events between many county pairs—complicated the decomposition process. Tensor decomposition is generally more robust when applied to datasets with fewer nodes and many events, such as public transportation data. As a result, the components generated by our decomposition often lacked clear interpretability, further underscoring the challenges of applying tensor decomposition to county-level migration patterns.

To accurately capture the multifaceted relationship between migration and health, future studies must also prioritize the development and integration of demographic-specific data sources. Current datasets overlook critical populations—such as undocumented migrants, low-income households, and young adults—whose migration patterns and health needs may significantly differ from the general population. These populations frequently experience unique health vulnerabilities and stressors related to migration, which are not reflected in generalized data. Tailored data sources that account for race, age, socio-economic status, and immigration status are essential for understanding the nuanced ways migration impacts health disparities across different demographic groups. Access to these demographic-specific data sources would allow researchers to capture a fuller, more accurate picture of how migration affects health outcomes, leading to more effective, equitable public health interventions.

Looking to the future, it is crucial to extend migration research to encompass the growing impacts of climate-related migration, as climate change increasingly disrupts environmental stability and livelihoods across the U.S. This type of migration is expected to intensify in the coming years, especially in areas vulnerable to extreme weather events, rising sea levels, and prolonged droughts. These shifts are likely to disproportionately affect already vulnerable populations, exacerbating health disparities in both origin and destination areas. Climate-induced migration may introduce new health challenges, including mental health stressors, exacerbated chronic diseases, and greater strain on healthcare resources in regions receiving large influxes of climate migrants. Applying tensor decomposition to understand these climate-related migration patterns could reveal critical insights into how climate change and migration jointly impact health outcomes and can help identify high-risk areas that would benefit most from targeted public health resources.

In conclusion, we explored the application of tensor decomposition to quantify the relationship between a county's position within a migration system and overall county health. While our findings were inconclusive, this study highlights the potential of using spatial and temporal analyses to better understand the complex interplay between migration and health outcomes. The exploratory nature of this work underscores the need for further research and methodological advancements to effectively capture and interpret these dynamics. Tailored public health interventions in regions experiencing significant migration, informed by a deeper understanding of migration systems and their health impacts, remain a critical priority. By considering both origin and destination dynamics, this study contributes to the broader conversation about how migration shapes public health in the United States.
